### 2020-07-27 generated/paper_EN.Rmd
Add preparated data xls

---
title:
  Concrete compressive strength prediction with machine learning
date: "April 22, 2020"
author: "Pedro Bernardino Alves Moreira"
indent: true
abstract:
 Compressive strength is the main characteristic of concrete. This work used a set of samples widely used to build 6 forecast models of this characteristic, each one specifically for an age, 3, 7, 14, 28, 56 and 100 days from 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers) and 6 relationships between the main ingredients (water/cement, coarse aggregate/cement, fine aggregate/cement, fine aggregate / coarse aggregate, water / coarse aggregate and water/fine aggregate) to make the predictions. The algorithm *Parallel Random Forest* was used (*parRF* of the *caret* package in *R*) . The results were satisfactory, being the best for the 3-day model (*RMSE* of 3.31) and the worst for the 100-day model (*RMSE* of 5.85). The 28-day model had *RMSE* of 4.72. The work indicated that the separation in a model for each age is promising and was able to generate compatible results or better than other studies, opening way for new models creation following this proposal, but using different algorithms or the combination of different algorithms.
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
bibliography: references.bib
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{supertabular}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
- \usepackage{caption}
- \usepackage{floatrow}
- \floatsetup[table]{capposition=top}
- \floatsetup[figure]{capposition=top}
- \captionsetup{options=chunk}
- \DeclareNewFloatType{chunk}{placement=H, fileext=chk, name=}
- \renewcommand{\thechunk}{\arabic{chunk}}
- \usepackage{indentfirst}
- \usepackage{sectsty} \sectionfont{\centering}
---
```{r setup, include=FALSE, cache=F}
# Install libraries, if not already installed
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(caret)) install.packages("caret")
if(!require(knitr)) install.packages("knitr")
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(gdata)) install.packages("gdata")
if(!require(dplyr)) install.packages("dplyr")
if(!require(tidyr)) install.packages("tidyr")
if(!require(cowplot)) install.packages("cowplot")
if(!require(pastecs)) install.packages("pastecs")
if(!require(factoextra)) install.packages("factoextra")
if(!require(reshape2)) install.packages("reshape2")
if(!require(purrr)) install.packages("purrr")
if(!require(gridExtra)) install.packages("gridExtra")
if(!require(questionr)) install.packages("questionr")
# Load libraries
library(tidyverse)
library(caret)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(gdata)
library(reshape2)
library(cowplot)
library(pastecs)
library(factoextra)
library(purrr)
library(gridExtra)
library(questionr)
# Set global options
options(pillar.sigfig = 5)
options(knitr.kable.NA = '')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo = TRUE, fig.align = 'center', cache=T, prompt=F, highlight=T )
oldSource <- knit_hooks$get("source")
knit_hooks$set(source = function(x, options) {
  x <- oldSource(x, options)
  x <- ifelse(!is.null(options$ref), paste0("\\label{", options$ref,"}", x), x)
  ifelse(!is.null(options$codecap), paste0("\\captionof{chunk}{", options$codecap,"}", x), x)
})
```
\newpage
\tableofcontents
\newpage
\listoffigures
\listoftables
\newpage

@@

## Materials
The methodology was carried out using RStudio software [@RStudio], an integrated virtual environment for code development in *R* [@RCore]. Throughout the process, all code executed was documented in the same order as its execution in [Appendix 3](#appendix3), and reference was always made to codes throughout the text. All relevant information related to the operating system and installed packages has been presented in [Appendix 1](#appendix1). In addition, an online and open-source repository was created in *Github*, housing all the code used to generate this work, the link was made available in [Appendix 2](#appendix2).

## Reproducibility
In order to guarantee reproducibility, whenever there was some code that could use probabilistic operations, a *seed* was defined before its execution, ensuring that when run on another machine, with the same version of *R* and the same *seed*, get the same result. The *seeds* can be checked throughout [Appendix 3](#appendix3) or directly on *Github*.

## Obtaining the data
The data was downloaded from the University of California Irvine website [@downloadData] (\ref{show-download-data}). In total there are 1030 samples with 9 columns. The samples were renamed and an id column was added to facilitate data manipulation (\ref{show-rename-dat-cols}). The columns were reordered to put the new id column in the first position (\ref{show-reorder-dat}). The first samples can be viewed in the table \ref{tab: first-samples}.

@@

```{r download-data, echo=F, include=F}
# Data download
url_base <- "https://archive.ics.uci.edu"
url <-  "/ml/machine-learning-databases/concrete/compressive/Concrete_Data.xls"
download.file(paste0(url_base, url), "data.xls")
dat <- read.xls("data.xls")
n_inicial_samples <- nrow(dat)
colnames(dat)
```
```{r rename-dat-cols, echo=F}
# Ranaming the columns
colnames(dat) <- c(
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
  )
dat$id <- seq.int(nrow(dat))
```
```{r reorder-dat, echo=F}
# Reordering the data
col_order <- c(
  "id",
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
)
dat <- dat[, col_order]
```
```{r col-names-and-units, echo=F}
# Defining column names and units
colNames <- c("ID", "Cement", "B.F.S.", "Fly ash", "Water",
             "Superp.", "C.Aggregate", "F.Aggregate", "Day", "Comp.Str.")
dfUnits <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
             "$kg/m^3$","$kg/m^3$","$kg/m^3$","","$MPa$")
```
```{r first-samples, echo=F}
# Table First samples
caption <- "First 6 samples"
kable(
    dat[1:6,],
    col.names = dfUnits,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

## Data preparation
The preparation of the data consisted of transforming the sample set in order to maintain only relevant data for the subsequent studies. Data that were considered irrelevant or that had the potential to add undesirable noise to the analysis were removed. In addition, the relevant data has been transformed to better fit the studies in the next steps.

### Initial data cleaning
Initially, there were 25 duplicate samples that were removed, resulting in a new total of 1005 samples (\ref{show-removing-duplicated-samples}).

@@

```{r removing-duplicated-samples, echo=F}
# Removing duplicated samples
n_distinct_samples <- dat %>% select(-c(id)) %>% n_distinct()
n_duplicated_samples <- n_inicial_samples - n_distinct_samples
dat <- dat[!duplicated(select(dat, -c(id))),]
n_samples <- nrow(dat)
```
The data show the variables in the columns and samples in the rows. However it was found that some samples are identical in proportions of ingredients, changing only the value of age and compressive strngth, for example, samples 653, 654, 678 and 681, shown in the table \ref{tab:similar-samples}.

@@

```{r similar-samples, echo=FALSE}
# Tabela - Samples with same composition
same_samples <- dat %>%
  filter(id %in% c(653, 678, 654, 681))
caption <- "Samples with same composition"
kable(
  same_samples[order(same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```
In addition, there are also samples with the same values and proportions of ingredients, but with different compressive strength, probably due to differences in the building process. This is the case, for example, of samples 472, 473 and 474, shown in the table  \ref{tab:similar-samples-2}.

@@

```{r similar-samples-2, echo=FALSE}
# Tabela - Same samples with different results
same_samples_2 <- dat %>%
  filter(id %in% c(472, 473, 474))
caption <- "Same samples with different results"
kable(
  same_samples_2[order(same_samples_2$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("scale_down"))
```
To facilitate the analysis of the samples, all samples that are the same in relation to the ingredients, have been assigned the same *id*. In addition, as the compressive strength at 28 days is the parameter to determine the grade of the concrete, only the elements containing that day among its samples were maintained. In the case of the same samples but with different results of compressive strength, the values were averaged. After all these changes (\ref{show-initial-data-cleaning}), the new total of samples was reduced to 970, containing 416 different settings for the proportions of ingredients.

@@

```{r similar-samples-same-id, echo=FALSE, include=TRUE}
# Tabela - Previous samples after processing
same_samples <- dat %>%
  filter(id == 653 | id == 472 & day == 28)
caption <- "Previous samples after processing"
kable(
  same_samples[order(same_samples$id, same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE,
  digits = 2
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

### Age selection
As previously described, the main age for analysis of compressive strength is 28 days, but other ages can also be used to build forecasting models. However, it is necessary to verify how relevant the data of these other ages are. Starting with the distribution of the samples in relation to each age  (\ref{show-boxplot}) shown in the figure \ref{fig:boxplot}.

@@

```{r pca-90-91-100, echo=F, include=T, fig.cap = cap, fig.height=3.5}
# Figure - Principal component analysis - 90, 91 e 100 days
dat_90_91_100 <- dat %>%
  ungroup() %>%
  filter(day %in% c(90, 91, 100)) %>%
  select(-c(id, mpa))
cap <- "Principal component analysis - 90, 91 e 100 days"
colnames(dat_90_91_100) <- c(
   "Cem.", "B.F.S.", "Fly.A.","Water","Sup.","C.Ag.","F.Ag.","day")
pca <- prcomp(select(dat_90_91_100, -c(day)), scale = TRUE)
habillage <- dat_90_91_100$day
fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage=habillage,
  addEllipses = TRUE,
  ellipse.level=0.75) +
  ggtitle("") +
  theme_bw() +
  coord_cartesian(xlim = c(-3, 3.5), ylim = c(3, -5))
```
Another important point considered, of the concrete nature itself, is the fact that the growth rate of its compressive strength decreases with time, reaching a certain stability value. The figure \ref{fig:mpa-on-time} shows the compressive strength over the days for samples with more than 5 data, that is, data available for at least 6 different ages (\ref{show-mpa-on-team}).

@@

```{r mpa-on-time, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Compressive strength through time
cap <- "Compressive strength through time"
ylabel <- "MPa"
xlabel <- "Age (days)"
dat_duplicated_only <- dat %>%
  group_by(id) %>%
  filter(n()>5) %>%
  select(id, mpa, day)
dat_duplicated_only %>%
  ggplot(aes(day, mpa, fill=id, alpha = 0.5)) +
  geom_line() +
  xlab(xlabel) +
  ylab(ylabel) +
  theme_bw() +
  theme(legend.position = "none")
```
For the reasons presented in the figures \ref{fig:boxplot}, \ref{fig:pca-90-91-100} and \ref{fig:mpa-on-time}, it was considered that the ages of 90, 91 and 100 days can be grouped to improve reading and decrease sample noise. They were converted to the same value, the age of 100 days was chosen (\ref{show-join-90-91-100}. As shown in the figure \ref{fig:mpa-on-time}, the resistance only increases, after 100 days the resistance to compression will be greater than or equal to the value of 90 or 91 days.

@@

```{r join-90-91-100, echo=F}
# Joining 90, 91 and 100 days data
ind_90 <- dat$id[which(dat$day == "90")]
ind_91 <- dat$id[which(dat$day == "91")]
ind_100 <- dat$id[which(dat$day == "100")]
sum(duplicated(c(ind_90, ind_91, ind_100))) # 0
dat <- dat %>%
  ungroup() %>%
  mutate(day = ifelse(day %in% c(91, 90), 100, day))
```
Another topic analyzed in the selection of ages was the observed frequency of each age value after this transformation from 90, 91 in 100 days, shown in the figure \ref{fig:freq-ages}. Some values of days have very low concentrations of samples, at the risk of damaging more than helping to create the models, so they were removed (\ref{show-remove-ages-lower-50}). The criterion adopted was to maintain only ages with a frequency greater than 50, only the values of 3, 7, 14, 28, 56 and 100 days.

@@

```{r freq-ages, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Ages frequency
cap <- "Ages frequency"
ylabel <- "Frequency"
xlabel <- "Age (days)"
dat %>%
  ggplot(aes(x = factor(day))) +
  geom_bar() +
  theme_bw() +
  xlab(xlabel) +
  ylab(ylabel)
```
```{r remove-ages-lower-50, echo=F}
# Removing ages with frequency lower than 50
dat <- dat[dat$day %in% c(3, 7, 14, 28, 56, 100),]
```

### Data reorganization
The samples were grouped to maintain only a separate sample from each set of configuration of the proportions of the ingredients, adding new variables/columns for resistance at each age (\ref{show-reorganizing-dat}). The result in the first samples after this processing is shown in the table \ref{tab:new-features}.

@@

```{r reorganizing-dat, echo=F}
# Data reorganization
dat <- dat %>%
  group_by_at(vars(-mpa)) %>%
  mutate(row_id = 1:n()) %>% ungroup() %>%
  spread(day, mpa, sep = "_") %>%
  select(-row_id)
```
```{r new-features, echo=F}
# Table - First 6 samples after reorganization
caption <- "First 6 samples after reorganization"
colNames2 = c("ID", "Cement", "B.F.S", "Fly ash", "Water",
             "Superp.", "Coarse Ag.", "Fine Ag.", "3 days", "7 days",
             "14 days", "28 days", "56 days", "100 days")
dfUnits2 <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
              "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$MPa$","$MPa$",
              "$MPa$","$MPa$","$MPa$","$MPa$")
kable(
    head(dat[order(dat$id),]),
    col.names = dfUnits2,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames2, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```
The number of samples and distinct samples after all this manipulation remained the same, a total of 416 (\ref{show-total-samples-2}).
```{r total-samples-2, echo=F}
# Total samples
n_samples <- nrow(dat) # 416
n_distinct_samples <- n_distinct(dat$id) # 416
```

>>>>

---
title:
  Concrete compressive strength prediction with machine learning
date: "April 22, 2020"
author: "Pedro Bernardino Alves Moreira"
indent: true
abstract: "Compressive strength is the main characteristic of concrete. The correct prediction of this parameter means cost and time reduction. This work built predictive models for 6 different ages of concrete samples (3, 7, 14, 28, 56, and 100 days). A set of data obtained in previous studies was used, a total of 1030 samples, with 9 variables: compressive strength, age, and 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers). Another 6 variables were added to represent the proportions of the main ingredients in each sample (water/cement, fine aggregate/cement, coarse aggregate/cement, fine aggregate/coarse aggregate, water/coarse aggregate, and water/fine aggregate). The predictive models were developed in *R* language, using the *caret* package with the *Parallel Random Forest* algorithm and repeated cross-validation technique to optimize the parameters. The results were satisfactory and compatible with other studies using the same data set. The most important model, 28 days old, obtained *RMSE* of 4.717. The 3-day model obtained the best result, *RMSE* of 3.310. The worst result was the 56-day model, with *RMSE* of 5.939. The work showed that the compressive strength of concrete can be predicted. The choice of creating a model for each age, instead of using age as a predictor, allowed to get compatible results with the available data at each age. It was a promising alternative since good results were achieved by training with just one algorithm. This work facilitates exploration and new efforts to predict the compressive strength of concrete, it can be replicated using different algorithms or the combination of several."
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
bibliography: references.bib
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{supertabular}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
- \usepackage{caption}
- \usepackage{floatrow}
- \floatsetup[table]{capposition=top}
- \floatsetup[figure]{capposition=top}
- \captionsetup{options=chunk}
- \DeclareNewFloatType{chunk}{placement=H, fileext=chk, name=}
- \renewcommand{\thechunk}{\arabic{chunk}}
- \usepackage{indentfirst}
- \usepackage{sectsty} \sectionfont{\centering}
---
```{r setup, include=FALSE, cache=F}
# Install libraries, if not already installed
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(caret)) install.packages("caret")
if(!require(knitr)) install.packages("knitr")
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(gdata)) install.packages("gdata")
if(!require(dplyr)) install.packages("dplyr")
if(!require(tidyr)) install.packages("tidyr")
if(!require(cowplot)) install.packages("cowplot")
if(!require(pastecs)) install.packages("pastecs")
if(!require(factoextra)) install.packages("factoextra")
if(!require(reshape2)) install.packages("reshape2")
if(!require(purrr)) install.packages("purrr")
if(!require(gridExtra)) install.packages("gridExtra")
if(!require(questionr)) install.packages("questionr")
if(!require(xlsx)) install.packages("xlsx")

# Load libraries
library(tidyverse)
library(caret)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(gdata)
library(reshape2)
library(cowplot)
library(pastecs)
library(factoextra)
library(purrr)
library(gridExtra)
library(questionr)
library("xlsx")

# Set global options
options(pillar.sigfig = 5)
options(knitr.kable.NA = '')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo = TRUE, fig.align = 'center', cache=T, prompt=F, highlight=T )
oldSource <- knit_hooks$get("source")
knit_hooks$set(source = function(x, options) {
  x <- oldSource(x, options)
  x <- ifelse(!is.null(options$ref), paste0("\\label{", options$ref,"}", x), x)
  ifelse(!is.null(options$codecap), paste0("\\captionof{chunk}{", options$codecap,"}", x), x)
})
```
\newpage
\tableofcontents
\newpage
\listoffigures
\listoftables
\newpage

@@

## Materials

The methodology was carried out using RStudio software [@RStudio], an integrated virtual environment for code development in *R* [@RCore]. Throughout the process, all code executed was documented in the same order as its execution in [Appendix 2](#appendix2), and reference was always made to codes throughout the text. All relevant information related to the operating system and installed packages has been presented in [Appendix 1](#appendix1).

## Reproducibility

In order to guarantee reproducibility, whenever there was some code that could use probabilistic operations, a *seed* was defined before its execution, ensuring that when run on another machine, with the same version of *R* and the same *seed*, get the same result. The *seeds* can be checked throughout [Appendix 2](#appendix2).

## Obtaining the data

The data was downloaded from the University of California Irvine website [@downloadData] (\ref{show-download-data}). In total there are 1030 samples with 9 columns. The samples were renamed and an id column was added to facilitate data manipulation (\ref{show-rename-dat-cols}). The columns were reordered to put the new id column in the first position (\ref{show-reorder-dat}). The first samples can be viewed in the table \ref{tab:first-samples}.

@@

```{r download-data, echo=F, include=F}
# Data download
url_base <- "https://archive.ics.uci.edu"
url <-  "/ml/machine-learning-databases/concrete/compressive/Concrete_Data.xls"
download.file(paste0(url_base, url), "data.xls")
dat <- read.xls("data.xls")
n_inicial_samples <- nrow(dat)
colnames(dat)
```

```{r rename-dat-cols, echo=F}
# Renaming the columns
colnames(dat) <- c(
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
  )
dat$id <- seq.int(nrow(dat))
```

```{r reorder-dat, echo=F}
# Reordering the data
col_order <- c(
  "id",
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
)
dat <- dat[, col_order]
```

```{r col-names-and-units, echo=F}
# Defining column names and units
colNames <- c("ID", "Cement", "B.F.S.", "Fly ash", "Water",
             "Superp.", "C.Aggregate", "F.Aggregate", "Day", "Comp.Str.")
dfUnits <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
             "$kg/m^3$","$kg/m^3$","$kg/m^3$","","$MPa$")
```

```{r first-samples, echo=F}
# Table - First samples
caption <- "First 6 samples"
kable(
    dat[1:6,],
    col.names = dfUnits,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

## Data preparation

The preparation of the data consisted of transforming the sample set in order to maintain only relevant data for the subsequent studies. Data that were considered irrelevant or that had the potential to add undesirable noise to the analysis were removed. In addition, the relevant data has been transformed to better fit the studies in the next steps.

### Initial data cleaning

Initially, there were 25 duplicate samples that were removed, resulting in a new total of 1005 samples (\ref{show-removing-duplicated-samples}).

@@

```{r removing-duplicated-samples, echo=F}
# Removing duplicated samples
n_distinct_samples <- dat %>% select(-c(id)) %>% n_distinct()
n_duplicated_samples <- n_inicial_samples - n_distinct_samples
dat <- dat[!duplicated(select(dat, -c(id))),]
n_samples <- nrow(dat)
```

The data show the variables in the columns and samples in the rows. However it was found that some samples are identical in proportions of ingredients, changing only the value of age and compressive strength, for example, samples 653, 654, 678 and 681, shown in the table \ref{tab:similar-samples}.

@@

```{r similar-samples, echo=FALSE}
# Table - Samples with same composition
same_samples <- dat %>%
  filter(id %in% c(653, 678, 654, 681))
caption <- "Samples with same composition"
kable(
  same_samples[order(same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

In addition, there are also samples with the same values and proportions of ingredients, but with different compressive strength, probably due to differences in the building process. This is the case, for example, of samples 472, 473 and 474, shown in the table  \ref{tab:similar-samples-2}.

@@

```{r similar-samples-2, echo=FALSE}
# Table - Same samples with different results
same_samples_2 <- dat %>%
  filter(id %in% c(472, 473, 474))
caption <- "Same samples with different results"
kable(
  same_samples_2[order(same_samples_2$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("scale_down"))
```

To facilitate the analysis of the samples, all samples that are the same in relation to the ingredients, have been assigned the same *id*. In addition, as the compressive strength at 28 days is the parameter to determine the grade of the concrete, only the elements containing that day among its samples were maintained. In the case of the same samples but with different results of compressive strength, the values were averaged. After all these changes (\ref{show-initial-data-cleaning}), the new total of samples was reduced to 970, containing 416 different settings for the proportions of ingredients.

@@

```{r similar-samples-same-id, echo=FALSE, include=TRUE}
# Table - Previous samples after processing
same_samples <- dat %>%
  filter(id == 653 | id == 472 & day == 28)
caption <- "Previous samples after processing"
kable(
  same_samples[order(same_samples$id, same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE,
  digits = 2
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

### Age selection

As previously described, the main age for analysis of compressive strength is 28 days, but other ages can also be used to build predictive models. However, it is necessary to verify how relevant the data of these other ages are. Starting with the distribution of the samples in relation to each age  (\ref{show-boxplot}) shown in the figure \ref{fig:boxplot}.

@@

```{r pca-90-91-100, echo=F, include=T, fig.cap = cap, fig.height=3.5}
# Figure - Principal component analysis - 90, 91 e 100 days
dat_90_91_100 <- dat %>%
  ungroup() %>%
  filter(day %in% c(90, 91, 100)) %>%
  select(-c(id, mpa))
cap <- "Principal component analysis - 90, 91 e 100 days"
colnames(dat_90_91_100) <- c(
   "Cem.", "B.F.S.", "Fly.A.","Water","Sup.","C.Ag.","F.Ag.","day")
pca <- prcomp(select(dat_90_91_100, -c(day)), scale = TRUE)
habillage <- dat_90_91_100$day
fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage=habillage,
  addEllipses = TRUE,
  ellipse.level=0.75) +
  ggtitle("") +
  theme_bw() +
  coord_cartesian(xlim = c(-3, 3.5), ylim = c(3, -5))
```

Another important point considered, of the concrete nature itself, is the fact that the growth rate of its compressive strength decreases with time, reaching a certain stability value. The figure \ref{fig:mpa-on-time} shows the compressive strength over the days for samples with more than 5 data, that is, data available for at least 6 different ages (\ref{show-mpa-on-time}).

@@

```{r mpa-on-time, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Compressive strength through time
cap <- "Compressive strength through time"
ylabel <- "MPa"
xlabel <- "Age (days)"
dat_duplicated_only <- dat %>%
  group_by(id) %>%
  filter(n()>5) %>%
  select(id, mpa, day)
dat_duplicated_only %>%
  ggplot(aes(day, mpa, fill=id, alpha = 0.5)) +
  geom_line() +
  xlab(xlabel) +
  ylab(ylabel) +
  theme_bw() +
  theme(legend.position = "none")
```

For the reasons presented in the figures \ref{fig:boxplot}, \ref{fig:pca-90-91-100} and \ref{fig:mpa-on-time}, it was considered that the ages of 90, 91 and 100 days can be grouped to improve reading and decrease sample noise. They were converted to the same value, the age of 100 days was chosen (\ref{show-join-90-91-100}). As shown in the figure \ref{fig:mpa-on-time}, the resistance only increases, after 100 days the resistance to compression will be greater than or equal to the value of 90 or 91 days.

@@

```{r join-90-91-100, echo=F}
# Joining 90, 91 and 100 days data
ind_90 <- dat$id[which(dat$day == "90")]
ind_91 <- dat$id[which(dat$day == "91")]
ind_100 <- dat$id[which(dat$day == "100")]
sum_duplicated <- sum(duplicated(c(ind_90, ind_91, ind_100))) # 0
dat <- dat %>%
  ungroup() %>%
  mutate(day = ifelse(day %in% c(91, 90), 100, day))
```

Another topic analyzed in the selection of ages was the observed frequency of each age value after this transformation from 90, 91 in 100 days, shown in the figure \ref{fig:freq-ages}. Some values of days have very low concentrations of samples, at the risk of damaging more than helping to create the models, so they were removed (\ref{show-remove-ages-lower-50}). The criterion adopted was to maintain only ages with a frequency greater than 50, only the values of 3, 7, 14, 28, 56 and 100 days.

@@

```{r freq-ages, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Ages frequency
cap <- "Ages frequency"
ylabel <- "Frequency"
xlabel <- "Age (days)"
dat %>%
  ggplot(aes(x = factor(day))) +
  geom_bar() +
  theme_bw() +
  xlab(xlabel) +
  ylab(ylabel)
```

```{r remove-ages-lower-50, echo=F}
# Removing ages with frequency lower than 50
dat <- dat[dat$day %in% c(3, 7, 14, 28, 56, 100),]
```

```{r dat-preparated, echo=F}
dat_preparated <- dat
```

The samples were grouped to maintain only one sample from each set of configuration of the proportions of the ingredients, adding new variables/columns for resistance at each age (\ref{show-reorganizing-dat}). The result in the first samples after this processing is shown in the table \ref{tab:new-features}.

@@

```{r reorganizing-dat, echo=F}
# Data reorganization
dat <- dat %>%
  group_by_at(vars(-mpa)) %>%
  mutate(row_id = 1:n()) %>% ungroup() %>%
  spread(day, mpa, sep = "_") %>%
  select(-row_id)
```

```{r new-features, echo=F}
# Table - First 6 samples after reorganization
caption <- "First 6 samples after reorganization"
colNames2 = c("ID", "Cement", "B.F.S", "Fly ash", "Water",
             "Superp.", "Coarse Ag.", "Fine Ag.", "3 days", "7 days",
             "14 days", "28 days", "56 days", "100 days")
dfUnits2 <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
              "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$MPa$","$MPa$",
              "$MPa$","$MPa$","$MPa$","$MPa$")
kable(
    head(dat[order(dat$id),]),
    col.names = dfUnits2,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames2, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

The number of samples and distinct samples after all this manipulation remained the same, a total of 416 (\ref{show-total-samples-2}).

```{r total-samples-2, echo=F}
# Total samples
n_samples <- nrow(dat) # 416
n_distinct_samples <- n_distinct(dat$id) # 416
```

### Adding new variables

To finish the data preparation, new columns were added to the dataset (\ref{show-new-features-2}). Starting with the concrete class, for example if the compressive strength is between 25 and 30, it receives the class *C25*. The inclusion of the class was important because the compressive strength in *MPa* is a continuous variable, which will be used in the regression models, but the class as a discrete variable can provide another visualization of the data. The approximate mix of concrete was also added, which represents the proportions of aggregates (fine and coarse) for cement. Other proportions between the main ingredients were also added. The new variables are presented in the table \ref{tab:new-features-table}.

```{r new-features-2, echo=F}
# Adding New features
concrete_class <- function(mpa){
  if (mpa >= 10) {
    s <- as.character(mpa)
    first <- substr(s, start = 1, stop = 1)
    second <- ifelse(substr(s, start = 2, stop = 2) >= 5, 5, 0)
  }
  else {
    first <- ""
    second <- "5"
  }
  paste("C", first, second, sep = "")
}
mix <- function(c, f_ag, c_ag){
  paste(
    1,
    round(f_ag/c, 0),
    round(c_ag/c, 0),
    sep = ":")
}
dat <- dat %>%
  mutate(class = sapply(day_28, concrete_class)) %>%
  mutate(class = as.factor(class)) %>%
  mutate(mix_app = factor(
    mix(cement, fine_aggregate, coarse_aggregate))) %>%
  mutate(`water_/_cement` = water / cement) %>%
  mutate(`fine_aggregate_/_cement` = fine_aggregate/cement) %>%
  mutate(`coarse_aggregate_/_cement` = coarse_aggregate/cement) %>%
  mutate(`fine_aggregate_/_coarse_aggregate` = fine_aggregate/coarse_aggregate) %>%
  mutate(`water_/_coarse_aggregate` = water/coarse_aggregate) %>%
  mutate(`water_/_fine_aggregate` = water/fine_aggregate)
lvl <- levels(dat$class)
dat$class <- factor(
  dat$class, 
  levels=c( "C5", sort(lvl[lvl!="C5"], decreasing=F)))
```

```{r new-features-table, echo=F}
# Table - New features
caption <- "New features"
colNames7 = c("ID", "Class","Approximated Mix",
             "Water / Cement", "Fine Ag. / Cement", 
             "Coarse Ag. / Cement", "Fine Ag. / Coarse Ag.",
             "Water / Coarse Ag.", "Water / Fine Ag.")
kable(
    head(dat[order(dat$id),][,c(1,15:22)]),
    col.names = colNames7,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c",
    digits = 4
    ) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  column_spec(2, width = "1.5cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4:9, width = "1.7cm")
```

```{r dat-preparated-2, echo=F}
dat_preparated_2 <- dat
```

```{r dat-preparated-to-xml, echo=F}
dat_preparated_2 <- dat
names(dat_preparated_2) <-  gsub("day_", "", names(dat_preparated_2))
dat_preparated_2 <- dat_preparated_2 %>%
  gather("age", "Mpa", "3":"100", na.rm = TRUE, convert = TRUE)

## Data visualization

In order to assess the need for further manipulation before building the models, in this step the 416 samples already processed were visualized and analyzed.

### Descriptive statistics

The table \ref{tab:stat-summ} presents the statistical data of the continuous variables (\ref {show-stat-summ}). The *Null* line represents the number of zeroed values for the ingredients, and the *NA* line represents the number of missing data. As the samples were filtered to maintain only sets of samples with known values of compressive strength at 28 days, the number of *NAs* is zero for that age. The figure \ref{fig:stat-summ-categorical} presents the statistical data of the discrete variables (\ref{show-stat-summ-categorical}).
